Constant Partying: \\ Introducing a Generic Toolkit for Recursive Partitioning in R

### 2014-03-18 pkg/partykit/inst/NUS-2014/constparty.Rnw
rename
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@470 255f4249-2263-4417-b1ea-8498779e27a7
nodeprune example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@476 255f4249-2263-4417-b1ea-8498779e27a7
added better pictures from M&S 1963 added party tree for M&S 1963 further touch-ups and improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@494 255f4249-2263-4417-b1ea-8498779e27a7

%% need no 
\usepackage{Sweave}
\definecolor{InputColor}{rgb}{0,0,0.3}
\definecolor{OutputColor}{rgb}{0.2,0.2,0.2}

%% commands
\let\code\texttt
\let\class\texttt
\let\fct\texttt
\let\doi\texttt
\newcommand{\dquote}[1]{``#1''}
\newcommand{\ui}{\underline{i}}
\newcommand{\oi}{\overline{\imath}}
\newcommand{\darrow}{\stackrel{\mbox{\tiny \textnormal{d}}}{\longrightarrow}}
\newcommand{\parrow}{\stackrel{\mbox{\tiny \textnormal{p}}}{\longrightarrow}}
\newcommand{\dotequals}{\stackrel{\cdot}{=}}
\newcommand{\efp}{\mathrm{\it efp}}
\newcommand{\given}{\, | \,}
\newcommand{\ltime}{\lambda_\mathrm{time}}
\newcommand{\lcomp}{\lambda_\mathrm{comp}}
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}

@@

<<pmml, eval=FALSE>>=
pm <- pmmlTreeModel(file.path(system.file("pmml", 
    package = "partykit"), "ttnc.pmml"))
@

@@

<<cache>>=
if(file.exists("slides-trees.rda")) load("slides-trees.rda") else {
<<rpart>>
<<ctree>>
<<evtree>>
<<pmml>>
<<mob>>
<<mytree>>
save(rp, ct, ev, pm, mb, myttnc, file = "slides-trees.rda")
}
@

@@

@@

\begin{center}
\includegraphics{aid}
\end{center}
JASA $\bold{58}(302)$ June 1963 (well, there were no early online versions, so
most people read the paper in 1964 for the first time). \\

@@

@@

\textbf{Challenge:} For implementing new algorithms in R, code
is required not only for fitting the tree model on the learning data
but also
\begin{itemize}
  \item representing fitted trees,
  \item printing trees,
  \item plotting trees,
  \item computing predictions from trees.
\end{itemize}

@@

\textbf{Design principles:} Toolkit for recursive partytitioning.
\begin{itemize}
  \item One `agnostic' base class which can encompass an extremely wide
        range of different types of trees.
  \item Subclasses for important types of trees, e.g., trees with constant
        fits in each terminal node.
  \item Nodes are recursive objects, i.e., a node can contain child nodes.
  \item Keep (learning) data out of the recursive node and split structure.
  \item Basic printing, plotting, and predicting for raw node structure.
  \item Customization via suitable panel or panel-generating functions.
  \item Coercion from existing objects (rpart, J48, etc.) to new class.
  \item Use simple/fast S3 classes and methods.
\end{itemize}

@@

    where \code{node} is a \class{partynode} and \code{data} the
    corresponding (learning) data (optionally without any rows) and
    \code{fitted} the corresponding fitted nodes.
\end{itemize}

@@

@@

\textbf{PMML:} Predictive Model Markup Language. XML-based format
exported by various software packages including SAS, SPSS, R/pmml.
Here, import QUEST tree.

@@

\begin{frame}[fragile]
\frametitle{mytree}

@@

@@

>>>>

%% instead of \usepackage{Sweave}
\definecolor{InputColor}{rgb}{0,0,0.3}
\definecolor{OutputColor}{rgb}{0.2,0.2,0.2}
\setkeys{Gin}{width=0.8\textwidth}
\RequirePackage{fancyvrb,upquote}
\renewcommand{\ttdefault}{laett}
\DefineVerbatimEnvironment{Scode}{Verbatim}{}
\DefineVerbatimEnvironment{Sinput}{Verbatim}{formatcom=\color{InputColor}}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{formatcom=\color{OutputColor}}
\newenvironment{Schunk}{\fontsize{9}{10}\selectfont}{}

%% commands
\newcommand{\code}[1]{{\tt \small #1}}
\let\class\code
\let\fct\code
\let\doi\code
\newcommand{\dquote}[1]{``#1''}
\newcommand{\ui}{\underline{i}}
\newcommand{\oi}{\overline{\imath}}
\newcommand{\darrow}{\stackrel{\mbox{\tiny \textnormal{d}}}{\longrightarrow}}
\newcommand{\parrow}{\stackrel{\mbox{\tiny \textnormal{p}}}{\longrightarrow}}
\newcommand{\dotequals}{\stackrel{\cdot}{=}}
\newcommand{\efp}{\mathrm{\it efp}}
\newcommand{\given}{\, | \,}
\newcommand{\ltime}{\lambda_\mathrm{time}}
\newcommand{\lcomp}{\lambda_\mathrm{comp}}
\newcommand{\argmin}{\operatorname{argmin}\displaylimits}

@@

<<pmml, eval=FALSE>>=
pm <- pmmlTreeModel(system.file("pmml", "ttnc.pmml",
    package = "partykit"))
@

@@

<<cache>>=
if(file.exists("constparty-trees.rda")) load("constparty-trees.rda") else {
<<rpart>>
<<ctree>>
<<evtree>>
<<pmml>>
<<mob>>
<<mytree>>
save(rp, ct, ev, pm, mb, myttnc, file = "constparty-trees.rda")
}
@

@@

<<Morgen+Sonquist-1963>>=
ms63d <- data.frame(
  earnings = numeric(),
  ethnicity = factor(, levels = 1:2, labels = c("other", "afam")),
  age = numeric(),
  occupation = factor(, levels = 1:2, labels = c("other", "farmer")),
  education = factor(, levels = 1:4, labels = c("elementary", "highschool", "college", "university"), ordered = TRUE)
)

ms63p <- partynode(1L,
  split = partysplit(2L, index = 1:2),
  kids = list(
    partynode(2L,
      split = partysplit(3L, breaks = 65, right = FALSE),
      kids = list(
        partynode(3L,
	  split = partysplit(4L, index = 1:2),
	  kids = list(
	    partynode(4L,
	      split = partysplit(5L, breaks = 3L, right = FALSE),
	      kids = list(
	        partynode(5L,
		  split = partysplit(5L, breaks = 2L, right = FALSE),
		  kids = list(
		    partynode(6L, info = "$5000"),
		    partynode(7L, info = "$6500")
		  )
		),
		partynode(8L,
		  split = partysplit(3L, breaks = 45, right = FALSE),
		  kids = list(
		    partynode(9L, info = "$7000"),
		    partynode(10L, info = "$9000")
		  )
		)
	      )
	    ),
	    partynode(11L, info = "$4500")
	  )
	),
	partynode(12L,
	  split = partysplit(5L, breaks = 2L, right = FALSE),
	  kids = list(
	    partynode(13L, info = "$2200"),
	    partynode(14L, info = "$4500")
	  )
	)
      ),
      info = "$5500"
    ),
    partynode(15L,
      split = partysplit(5L, breaks = 2L, right = FALSE),
      kids = list(
        partynode(16L, info = "$2500"),
	partynode(17L, info = "$5000")
      )
    )
  ),
  info = "$5000"
)
ms63p <- party(ms63p, ms63d)
@

@@

\begin{center}
\includegraphics{aid-abstract}
\end{center}
JASA $\bold{58}(302)$ June 1963 (well, there were no early online versions, so
most people read the paper in 1964 for the first time). \\

@@

\vspace*{-0.5cm}
\begin{center}
\includegraphics[width=\textwidth]{aid-tree}
\end{center}

@@

\textbf{Challenge:} For implementing new algorithms in R, code
is required not only for fitting the tree model
but also
\begin{itemize}
  \item representing fitted trees,
  \item printing trees,
  \item plotting trees,
  \item computing predictions from trees.
\end{itemize}

@@

\textbf{Design principles:} Toolkit for recursive partytitioning.
\begin{itemize}
  \item One `agnostic' base class which can encompass an extremely wide
        range of different types of trees.
  \item Subclasses for important types of trees, e.g., trees with constant
        fits in each terminal node.
  \item Nodes are recursive objects: nodes can contain child nodes.
  \item Keep data out of the recursive node and split structure.
  \item Basic print, plot, and predict for raw node structure.
  \item Customization via panel or panel-generating functions.
  \item Coercion from existing objects (e.g., rpart) to new class.
  \item Use simple/fast S3 classes and methods.
\end{itemize}

@@

    where \code{node} is a \class{partynode} and \code{data} the
    corresponding (learning) data (optionally without any rows) and
    \code{fitted} the corresponding fitted node ids.
\end{itemize}

@@

<<ms63-party, echo=TRUE, results=verbatim>>=
ms63d
pn <- partynode(1L,
  split = partysplit(2L, index = 1:2),
  kids = list(
    partynode(2L, info = "$5500"),
    partynode(3L, info = "$3000")
  )
)
py <- party(pn, ms63d)
print(py)
plot(py)
@

<<ms63-plot1, fig=TRUE>>=
plot(py)
@

\setkeys{Gin}{width=1.15\textwidth}
<<ms63-plot2, fig=TRUE, height=8, width=14>>=
plot(ms63p, drop = TRUE)
@

@@

\textbf{PMML:} Predictive Model Markup Language. XML-based format
exported by various software packages including SAS, SPSS, R/pmml.
Here, import QUEST tree from SPSS.

@@

\begin{frame}
\frametitle{mytree}

@@

\begin{frame}[fragile]
\frametitle{OK, fire away!}

@@

\begin{frame}[fragile]
\frametitle{Too big a tree? Prune!}

\setkeys{Gin}{width=\textwidth}
\hspace*{-0.4cm}%
<<Titanic-mytree-prune, echo = TRUE, fig = TRUE, height=8, width=12>>=
plot(nodeprune(myttnc, 10))
@
